Probabilistic forecasting of hourly Emergency Department arrivals

### 2022-12-28 paper/paper.Rmd
added missing references

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
[@derlet2002overcrowding]. Also, it is related to increasing length of
stay [@muhammet2015forecasting], low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records and other [@rostami2020anticipating].

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average
(ARIMA) [@hyndman2021forecasting], Autoregressive Conditional
Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive
model [@lutkepohl2013vector], TBATS [@de2011forecasting] and Artificial
Neural Networks [@hyndman2021forecasting]. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow _optimal_ decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

The variables _1--3_ allow modelling the seasonal patterns on the
appropriate level of detail throughout the year, while _4_ covers the
changes in admittance due to calendar events. Finally, _5_ is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g. people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \@ref(eq:PoissonRegression) introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
`alm()` function from the `greybox` package [@Svetunkov2021Greybox] for
R [@RTeam2021] for the experiments and denote this model as "Poisson
Regression".

@@

Gradient Boosting Machines (GBM) are a tree-based machine learning model
for regression and classification. Here we produce probabilistic
forecasts of ED arrivals via multiple quantile regression using GBMs as
implemented in the R package `gbm` [@Greenwell2020]. GBMs are a
best-in-class algorithm for similar regression problems characterised by
modest volumes of training data and possible interactions between input
features, and are therefore an appealing choice for ED arrival
forecasting. Here GBMs are fit with the following features: hour of the
day, day of the week, school holiday, temperature and day of the year.
Hyperparameters are chosen via grid search on the training data, and
were chosen as follows: 500 trees, an interaction depth of 2, and a
shrinkage 0.1. An advantage of GBMs is their ability to learn
interactions in comparison to the additive models presented later which
require the user to specify possible interactions between inputs. Here
GBM is included as a reference for the performance of \`out-of-the-box'
machine learning models applied with minimal effort. We also explored
the possibility of performing linear regression with additive models,
but these did not perform as well as GBM so are omited for brevity.

@@

To compare model performance, and the significance of any apparent
difference in performance, we will use skill scores, which can be
calculated for any metric via: \begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} (\#eq:skillscore)
\end{equation} where $M$ is the metric's value for the method being
considered, $M_\mathrm{ref}$ is the metric's value for a reference
method. The skill score shows by how many per cent the reference
approach is worse than the one under consideration. We will use
bootstrap re-sampling of skill scores to determine if the differences in
forecast performance (i.e. positive or negative skill) are significantly
different from zero [@Efron1981Bootstrap]. Here we use the best
performing simple benchmark, Naive (explained in Subsection
\@ref(naive)), as the reference model and employ a block-bootstrap
with blocks of length 24h to account for temporal correlations of the
underlying data [@hongyi1996bootstrapping; @Bergmeir2016303].

@@

\textcolor{blue}{
The dataset used in this study does not include the COVID-19 pandemic. During COVID-19, the dynamic of ED has changed. That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period. However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}

>>>>

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
[@derlet2002overcrowding]. Also, it is related to increasing length of
stay [@muhammet2015forecasting], low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records and other [@ROSTAMITABAR20221197].

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average
(ARIMA) [@hyndman2021forecasting], Autoregressive Conditional
Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive
model [@lutkepohl2013vector], TBATS [@de2011forecasting] and Artificial
Neural Networks [@hyndman2021forecasting]. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient [@wright2006reexamining]. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently [@ramos2013probabilistic;@ROSTAMITABAR20221197]. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow _optimal_ decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

The variables _1--3_ allow to model the seasonal patterns on the
appropriate level of detail throughout the year, while _4_ covers the
changes in admittance due to calendar events. Finally, _5_ is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g. people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \@ref(eq:PoissonRegression) introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
`alm()` function from the `greybox` package [@Svetunkov2021Greybox] for
R [@RTeam2021] for the experiments and denote this model as "Poisson
Regression".

@@

Gradient Boosting Machines (GBM) are a tree-based machine learning model
for regression and classification. Here we produce probabilistic
forecasts of ED arrivals via multiple quantile regression using GBMs as
implemented in the R package `gbm` [@Greenwell2020]. GBMs are a
best-in-class algorithm for similar regression problems characterised by
modest volumes of training data and possible interactions between input
features, and are therefore an appealing choice for ED arrival
forecasting. Here GBMs are fit with the following features: hour of the
day, day of the week, school holiday, temperature and day of the year.
Hyperparameters are chosen via grid search on the training data, and
were chosen as follows: 500 trees, an interaction depth of 2, and a
shrinkage 0.1. An advantage of GBMs is their ability to learn
interactions in comparison to the additive models presented later which
require the user to specify possible interactions between inputs. Here, GBM is included as a reference for the performance of _out-of-the-box_
machine learning models applied with minimal effort [@ridgeway2007generalized]. We also explored
the possibility of performing linear regression with additive models,
but these did not perform as well as GBM so are omited for brevity.

@@

To compare model performance, and the significance of any apparent
difference in performance, we will use skill scores, which can be
calculated for any metric via: \begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} (\#eq:skillscore)
\end{equation} where $M$ is the metric's value for the method being
considered, $M_\mathrm{ref}$ is the metric's value for a reference
method. The skill score shows by how many percent the reference
approach is worse than the one under consideration. We will use
bootstrap re-sampling of skill scores to determine if the differences in
forecast performance (i.e. positive or negative skill) are significantly
different from zero [@Efron1981Bootstrap]. Here we use the best
performing simple benchmark, Naive (explained in Subsection
\@ref(naive)), as the reference model and employ a block-bootstrap
with blocks of length 24h to account for temporal correlations of the
underlying data [@hongyi1996bootstrapping; @Bergmeir2016303].

@@

\textcolor{blue}{
The dataset used in this study does not include the period of the COVID-19 pandemic. During COVID-19, the dynamic of ED has changed. That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period. However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}

### 2022-12-28 paper/paper.tex
added missing references

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
\citep{derlet2002overcrowding}. Also, it is related to increasing length of
stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records and other \citep{rostami2020anticipating}.

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
Neural Networks \citep{hyndman2021forecasting}. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

Gradient Boosting Machines (GBM) are a tree-based machine learning model
for regression and classification. Here we produce probabilistic
forecasts of ED arrivals via multiple quantile regression using GBMs as
implemented in the R package \texttt{gbm} \citep{Greenwell2020}. GBMs are a
best-in-class algorithm for similar regression problems characterised by
modest volumes of training data and possible interactions between input
features, and are therefore an appealing choice for ED arrival
forecasting. Here GBMs are fit with the following features: hour of the
day, day of the week, school holiday, temperature and day of the year.
Hyperparameters are chosen via grid search on the training data, and
were chosen as follows: 500 trees, an interaction depth of 2, and a
shrinkage 0.1. An advantage of GBMs is their ability to learn
interactions in comparison to the additive models presented later which
require the user to specify possible interactions between inputs. Here
GBM is included as a reference for the performance of `out-of-the-box'
machine learning models applied with minimal effort. We also explored
the possibility of performing linear regression with additive models,
but these did not perform as well as GBM so are omited for brevity.

@@

To compare model performance, and the significance of any apparent
difference in performance, we will use skill scores, which can be
calculated for any metric via: \begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} \label{eq:skillscore}
\end{equation} where \(M\) is the metric's value for the method being
considered, \(M_\mathrm{ref}\) is the metric's value for a reference
method. The skill score shows by how many per cent the reference
approach is worse than the one under consideration. We will use
bootstrap re-sampling of skill scores to determine if the differences in
forecast performance (i.e.~positive or negative skill) are significantly
different from zero \citep{Efron1981Bootstrap}. Here we use the best
performing simple benchmark, Naive (explained in Subsection
\ref{naive}), as the reference model and employ a block-bootstrap
with blocks of length 24h to account for temporal correlations of the
underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.

>>>>

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
\citep{derlet2002overcrowding}. Also, it is related to increasing length of
stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records and other \citep{ROSTAMITABAR20221197}.

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
Neural Networks \citep{hyndman2021forecasting}. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently \citep{ramos2013probabilistic, ROSTAMITABAR20221197}. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

Gradient Boosting Machines (GBM) are a tree-based machine learning model
for regression and classification. Here we produce probabilistic
forecasts of ED arrivals via multiple quantile regression using GBMs as
implemented in the R package \texttt{gbm} \citep{Greenwell2020}. GBMs are a
best-in-class algorithm for similar regression problems characterised by
modest volumes of training data and possible interactions between input
features, and are therefore an appealing choice for ED arrival
forecasting. Here GBMs are fit with the following features: hour of the
day, day of the week, school holiday, temperature and day of the year.
Hyperparameters are chosen via grid search on the training data, and
were chosen as follows: 500 trees, an interaction depth of 2, and a
shrinkage 0.1. An advantage of GBMs is their ability to learn
interactions in comparison to the additive models presented later which
require the user to specify possible interactions between inputs. Here, GBM is included as a reference for the performance of \emph{out-of-the-box}
machine learning models applied with minimal effort \citep{ridgeway2007generalized}. We also explored
the possibility of performing linear regression with additive models,
but these did not perform as well as GBM so are omited for brevity.

@@

To compare model performance, and the significance of any apparent
difference in performance, we will use skill scores, which can be
calculated for any metric via: \begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} \label{eq:skillscore}
\end{equation} where \(M\) is the metric's value for the method being
considered, \(M_\mathrm{ref}\) is the metric's value for a reference
method. The skill score shows by how many percent the reference
approach is worse than the one under consideration. We will use
bootstrap re-sampling of skill scores to determine if the differences in
forecast performance (i.e.~positive or negative skill) are significantly
different from zero \citep{Efron1981Bootstrap}. Here we use the best
performing simple benchmark, Naive (explained in Subsection
\ref{naive}), as the reference model and employ a block-bootstrap
with blocks of length 24h to account for temporal correlations of the
underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.
